Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-12-27 paper/paper.tex
tweaks merging Alistair edits

The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\citep{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
As an example of the former case, expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only expressed viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS, the viral gene segment encoding the major interferon antagonist, NS1.

@@

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which, as mentioned, is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.

@@

These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including the probability that a cell triggers an IFN response.
We identified four types of defects in single IFN+ cells that we validated increase IFN induction. 
Two types of defects---absence of NS and mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different types of viral defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

But although variation within viral populations is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
Additionally, none of the viral defects in IFN+ cells induce IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\citep{oneal2018west}.

>>>>

The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).

The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells also lacked NS or had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
As an example of the former case, expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only expressed viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.

@@

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
%The NS gene encodes influenza's primary innate-immune antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.

@@

These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including the probability that a cell triggers an IFN response.
We identified four types of defects in single IFN+ cells that we validated increase IFN induction. 
Two types of defects---absence of NS and mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different defects also likely varies across conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

But although variation in viral populations is clearly important, it only partially explains heterogeneity among influenza-infected cells.
We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
Of course, our single-cell virus sequencing only calls mutations present in a cell at high frequency.
However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\citep{oneal2018west}.
